Researchers' perceptions of malaria eradication: findings from a mixed-methods analysis of a large online survey

### 2018-04-24 paper.tex
final for submission

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Researchers' perceptions of malaria eradication: findings from a mixed-methodsanalysis of a large online survey},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Researchers' perceptions of malaria eradication: findings from a
mixed-methodsanalysis of a large online survey}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

@@

\begin{center}
\begin{large}

Joe Brew\footremember{isglobal}{Barcelona Institute for Global Health: c/ Rosselló, 132, 5è 2a. 08036, Barcelona, Spain}\footremember{cism}{Centro de Investigação em Saúde de Manhiça: Vila da Manhiça, Bairro Cambeve, Rua 12, Distrito da Manhiça, CP 1929, Maputo, Mozambique}\footremember{vu}{VU University Amsterdam: De Boelelaan 1105, 1081 HV Amsterdam, Netherlands}, Menno Pradhan\footrecall{vu}\footremember{uva}{University of Amsterdam: REC E, Roetersstraat 11, Amsterdam, Netherlands}, Jacqueline Broerse\footrecall{vu}, Elisa Sicuri\footrecall{isglobal}\footremember{icl}{Imperial College London: South Kensington Campus, London SW7 2AZ, U.K.}\footrecall{cism}

\end{large}
\end{center}

@@

Quantifying an event's probability and time frame is essential for calculating its expected value. Uuncertainty regarding feasibility makes it difficult for policymakers and public health practitioners to forecast the return on investment of malaria eradication. Since the opportunity cost of investments in eradication-specific interventions can be high - particularly in contexts with other urgent health priorities - understanding eradication's expected value is essential. In a large online survey of malaria researchers (n = 844), we query perceptions regarding the likelihood and time frame of eradication. We assess pessimism/optimism (via the proxy of years-to-eradication), and adjust for selection bias. We also carry out a qualitative analysis of the free-form comments provided by 61\% of respondents. Our results show that researchers are skeptical of the feasibility of near-term malaria eradication, with most believing that eradication will not be achieved for 50 years or more.
\end{changemargin}
\end{center}

\noindent\fbox{%
    \parbox{\textwidth}{%
        \subsection*{Research Highlights}
        \begin{itemize}
          \item Employs a very large online survey of specialist researchers
          \item Assesses perceived time frame and likelihood of global malaria eradication among researchers
          \item Results suggest that most malaria researchers believe short-term global malaria eradication to be unlikely
          \item Identifies challenges pertaining to eradication through qualitative analysis
        \end{itemize}
      \subsection*{Keywords}
        Malaria, eradication, elimination, mixed methods, survey, crowdsourcing 
    }%
}

@@

@@

\caption{Countries with malaria: Observed and WHO objectives}\label{fig:unnamed-chunk-4}
\end{figure}

@@

Proponents of disease eradication point to the success of historical and
current campaigns (smallpox and polio, respectively), and highlight the
long term benefits in health and wealth to future generations (Barrett,
2013). One often cited figure is for smallpox eradication, in which the
investment of each dollar led to a return of greater than 400 (Barrett,
2006); unlike smallpox (for which analysis was able to be
retrospectively constructed based on observed costs and benefits), in
the case of a prevalent disease, forecasting models have to be used in
order to prospectively estimate the value of an investment.
\st{Additionally the estimated impact of malaria on economic indicators is heterogeneous, with singificant impact in areas of Sub-Saharan Africa - where the endemicity is high - and little to no impact in areas of low endemicity}
(Cutler et al., 2010) (Barofsky et al., 2015).

The general objective of eradication serves to inspire, rally funder
support, motivate researchers, and focus the efforts of public health
practitioners. But to the extent that malaria eradication (by
definition) has never occurred, the parameters needed for a ex-post
cost-benefit analysis are unknown. Nonetheless, prospectively estimating
eradication's return on investment is crucial to deciding when and how
to pursue the goal, especially in light of the high direct and
opportunity costs of eradication-specific interventions. One approach to
cost-benefit analysis is the use of mathematical and stochastic models.
These have been applied to diseases which are closer to eradication than
malaria, since the uncertainty around model input parameters is less,
requiring fewer cascading assumptions in order to present possible
comparative scenarios. For example, Kastner et al. were able to describe
4 relatively realistic pathways to Lympathic Filariasis eradication, as
well as the pre-requisite role and magnitude of certain interventions
(Kastner et al., 2015). A similar modeling framework was then used to
estimate the cost of eradication (Kastner et al., 2017). A recent
modeling analysis on onchocerciasis eradication, based on a disease
transmission model, showed that the costs of elimination (relative to
staying in ``control mode'') in Africa would be far lower even in the
short term, thanks to the improvements it would lead to in both
treatment times and prevented surveillance costs. Given the relative
proximity of eradication, and the narrow geographic scope, the authors
were also able to estimate the timeline to eradication (Kim et al.,
2017).

This level of detail and specificity in modeling the cost-benefit
landscape of malaria eradication, unfortunately, is not possible (or at
least not commonly undertaken), given its high prevalence and
epidemiological complexity. In fact, there are no mathematical models
(to the authors' knowledge) estimating the likelihood and time frame of
eradication, or its derivative cost-benefit ratio. While a great deal of
rigorous studies exist forecasting the effect of certain tools and
strategies relative to alternatives (Brady et al., 2017) or aimed at
regional disease elimination (Kim et al., 2015), these analyses do not
endeavor to estimate the generalization of those tools or strategies to
other areas. Globally, where mathematical modeling has been used for the
purposes of forecasting the future epidemiology of malaria, the methods
have generally been aimed at optimizing elimination methods (Slater et
al., 2015), determining whether a strategy is scalable (Brady et al.,
2017), guiding funding and drug development (Patouillard et al., 2017)
(Slater et al., 2017), or comparing a range of hypothetical morbidity
scenarios (Winskill et al., 2017), rather than assessing the likelihood
of or time until the occurrence of those scenarios. In other words,
researchers are (rightfully) focused on the \emph{how} of eradication,
rather than the forecasting of its timeframe or likelihood, the lack of
attention paid to the latter likely being a reflection of its
complexity, not lack of importance.

To the extent that (i) estimating eradication's likelihood and timeframe
is essential to forecasting the cost-benefit ratio of eradication
interventions, and (ii) mathematical models generally do not venture
into areas of such high uncertainty, alternative methods are needed.
Just as the stock market aggregates perceptions to provide a fairly
accurate assessment of something as complex as a company's value,
aggregating perceptions may be a useful tool for tackling the complexity
of malaria eradication. Many studies have shown value in expert
elicitation as a means to reduce uncertainty and inform decision making
(Morgan, 2014).
\st{But experts can sometimes be mistaken, a fact implicitly recognized when patients request a "second opinion" from another doctor when facing uncertainty. Additionally, relying on only a small panel of experts for a problem as complex as malaria exposes one to bias, just as an overly concentrated investment portfolio exposes one to more risk.}
As Sir Francis Galton demonstrated in his famous study in which he
showed that the crowds' aggregated estimates of cow's weight formed a
quasi-normal distribution centered around the true weight (Wallis, 2014)
(Galton, 1907), averaging the perceptions of many can be more accurate
than taking the opinion of any single expert, since the biases of
diverse viewpoints can be complementary and symbiotic. Measuring
consensus and discord among disease-specific researchers from a variety
of disciplines can serve as a barometer of (informed) opinion, both
guiding resources and identifying areas of concern (Keenan et al.,
2013).

@@

@@

\st{A great deal of previous research from serveral endemic areas already covers the cost per case prevented [@Sicuri2011] [@Silumbe2015] [@Btto2016] [@IlungaIlunga2014] [@Dalaba2015]. Likewise, literature exists which could serve as a model for quantifying the location-specific opportunity costs associated with funneling funds towards malaria eradication [@Stuckey2014] [@White2011] [@Korenromp2012]. The correct discount rate for estimating the value of future lives saved is more of a philosophical question than an economic one. This leaves only the probability and time-frame to eradication, questions which have been addressed anecdotally, but never answered quantitatively.}

\st{Though the case-specific marginal cost of eradication can be expected to be high (relative to a simple control approach), successful eradication would mean massive recurring savings in the long-term [@Barrett2013]. However, to the extent that the case-specific marginal cost of prevention in an eradication campaign is high, estimating the likelihood of success is fundamental to the correct distribution of resources, particularly in low-income environments. }

@@

{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-5-1}

@@

\caption{Perceptions of likelihood of eradication}\label{fig:unnamed-chunk-5}
\end{figure}

@@

{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-11-1}

@@

\caption{Aggregated perceived likelihood of eradication over time, adjusted for sample selection and unadjusted.}\label{fig:unnamed-chunk-11}
\end{figure}

@@

Comments were also classified as descriptive or prescriptive. A majority
(59.26\%) were descriptive. Descriptive commenters were more pessimistic
(in regards to perceived years until eradication) than prescriptive
commenters, though this difference did not reach the level of
statistical significance (p = 0.21). Descriptive comments also received
sentiment polarity scores which were more negative than prescriptive
comments, although again this difference did not reach the level of
statistical significance (p = 0.18).

@@

This study included the first authors of indexed journals. Though
certainly a group with important knowledge related to malaria, this
misses malaria control program employees, health agency workers, and
other stakeholders. Their experiences and viewpoints may be different
from those of academics, and arguably more relevant. Additionally, in
public health journals, first authors are often more junior than their
collaborators. By focusing solely on first authors, we may have
unintentionally created a sample which is qualitatively different than
the true universe of malaria researchers; to the extent that in our
result pool researchers with less experience (as represented through
publications) tended to be more ``optimistic'' regarding eradication, it
is reasonable to assume that the restriction of first authors may have
lead to an overly optimistic sample, making the results of the survey
even more striking.

@@

Our survey attempted to quantify the likelihood of, and time frame to,
malaria eradication by gauging the collective opinion of malaria
researchers. The findings of our survey show researchers expressing
hesitance about the likelihood of eradication and suggesting a long time
frame until it is achieved. The causes for scepticism are diverse, but
common themes were the need for innovation, systemic challenges, and the
complexity of the disease and its transmission.

@@

@@

\hypertarget{ref-Kim2015}{}
Kim, Y.E., Sicuri, E., Tediosi, F., 2015. Financial and economic costs
of the elimination and eradication of onchocerciasis (river blindness)
in africa. PLOS Neglected Tropical Diseases 9, e0004056.
\url{https://doi.org/10.1371/journal.pntd.0004056}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Researchers' perceptions of malaria eradication: findings from a mixed-methods analysis of a large online survey},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Researchers' perceptions of malaria eradication: findings from a
mixed-methods analysis of a large online survey}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

\usepackage{setspace}
\doublespacing

\usepackage[right, modulo]{lineno}
\linenumbers

@@

@@

Quantifying an event's probability and time frame is essential for calculating its expected value. Uncertainty regarding feasibility makes it difficult for policymakers and public health practitioners to forecast the return on investment of malaria eradication. Since the opportunity cost of investments in eradication-specific interventions can be high - particularly in contexts with other urgent health priorities - understanding eradication's expected value is essential. In a large online survey of malaria researchers (n = 844), we query perceptions regarding the likelihood and time frame of eradication. We assess perceived years to and likelihood of eradication, and adjust for selection bias. We also carry out a qualitative analysis of the free-form comments provided by 61\% of respondents. Our results show that researchers are skeptical of the feasibility of near-term malaria eradication, with most believing that eradication will not be achieved for 50 years or more.
\end{changemargin}
\end{center}

\vspace{10mm}

\noindent\fbox{%
    \parbox{\textwidth}{%
        \subsection*{Research Highlights}
        \begin{itemize}
          \item Employs a very large online survey of specialist researchers
          \item Assesses perceived time frame and likelihood of global malaria eradication among researchers
          \item Results suggest that most malaria researchers believe short-term global malaria eradication to be unlikely
          \item Identifies challenges pertaining to eradication through qualitative analysis
        \end{itemize}
      \subsection*{Keywords}
        Malaria; eradication; elimination; mixed methods; survey; crowdsourcing 
    }%
}

\newpage

@@

{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-3-1}

@@

\caption{Countries with malaria: Observed and WHO objectives}\label{fig:unnamed-chunk-3}
\end{figure}

@@

Proponents of disease eradication point to the success of historical and
current campaigns (smallpox and polio, respectively), and highlight the
long term benefits in health and wealth to future generations (Barrett,
2013). One often cited figure is for smallpox eradication, in which the
investment of each dollar led to a return of greater than 400 (Barrett,
2006); unlike smallpox (for which analysis was able to be
retrospectively constructed based on observed costs and benefits), in
the case of a prevalent disease, forecasting models have to be used in
order to prospectively estimate the value of an investment. Additionally
the estimated impact of malaria on economic indicators is heterogeneous,
with singificant impact in areas of Sub-Saharan Africa - where the
endemicity is high - and little to no impact in areas of low endemicity
(Cutler et al., 2010) (Barofsky et al., 2015).

The general objective of eradication serves to inspire, rally funder
support, motivate researchers, and focus the efforts of public health
practitioners. The recently established Lancet Commission on malaria
eradication states that ``eradication is the only acceptable end goal''
(while acknowledging that there exists ``scepticism about {[}its{]}
technical feasibility'') (Chen et al., 2018). To the extent that malaria
eradication (by definition) has never occurred, the parameters needed
for a ex-post cost-benefit analysis are unknown. Nonetheless,
prospectively estimating eradication's return on investment is crucial
to deciding when and how to pursue the goal, especially in light of the
high direct and opportunity costs of eradication-specific interventions.
One approach to economic evaluation is the use of mathematical and
stochastic models. These have been applied to diseases which are closer
to eradication than malaria, since the uncertainty around model input
parameters is less, requiring fewer cascading assumptions in order to
present possible comparative scenarios. For example, Kastner et al. were
able to describe 4 relatively realistic pathways to Lympathic Filariasis
eradication, as well as the pre-requisite role and magnitude of certain
interventions (Kastner et al., 2015). A similar modeling framework was
then used to estimate the cost of eradication (Kastner et al., 2017). A
recent modeling analysis on onchocerciasis eradication, based on a
disease transmission model, showed that the costs of elimination
(relative to staying in ``control mode'') in Africa would be far lower
even in the short term, thanks to the improvements it would lead to in
both treatment times and prevented surveillance costs. Given the
relative proximity of eradication, and the narrow geographic scope, the
authors were also able to estimate the timeline to eradication (Kim et
al., 2017).

This level of detail and specificity in the economic evaluation
landscape of malaria eradication, unfortunately, is not possible (or at
least not commonly undertaken), given its high prevalence and
epidemiological complexity. In fact, there are no mathematical models
(to the authors' knowledge) estimating the likelihood and time frame of
eradication, or its derivative cost-benefit ratio. Globally, where
mathematical modeling has been used for the purposes of forecasting the
future epidemiology of malaria, the methods have generally been aimed at
optimizing elimination methods (Slater et al., 2015), determining
whether a strategy is scalable (Brady et al., 2017), guiding funding and
drug development (Patouillard et al., 2017) (Slater et al., 2017), or
comparing a range of hypothetical morbidity scenarios (Winskill et al.,
2017), rather than assessing the likelihood of or time until the
occurrence of those scenarios.

To the extent that (i) estimating eradication's likelihood and timeframe
is essential to forecasting the cost-benefit ratio of eradication
interventions, and (ii) mathematical models generally do not venture
into areas of such high uncertainty, alternative methods are needed.
Just as the stock market aggregates perceptions to provide a fairly
accurate assessment of something as complex as a company's value,
aggregating perceptions may be a useful tool for tackling the complexity
of malaria eradication. Many studies have shown value in expert
elicitation as a means to reduce uncertainty and inform decision making
(Morgan, 2014). As Francis Galton demonstrated in his famous study in
which he showed that the crowds' aggregated estimates of cow's weight
formed a quasi-normal distribution centered around the true weight
(Wallis, 2014) (Galton, 1907), averaging the perceptions of many can be
more accurate than taking the opinion of any single expert, since the
biases of diverse viewpoints can be complementary and symbiotic.
Measuring consensus and discord among disease-specific researchers from
a variety of disciplines can serve as a barometer of (informed) opinion,
both guiding resources and identifying areas of concern (Keenan et al.,
2013).

@@

Though the case-specific marginal cost of eradication can be expected to
be high (relative to a simple control approach), successful eradication
would mean massive recurring savings in the long-term (Barrett, 2013).
However, to the extent that the case-specific marginal cost of
prevention in an eradication campaign is high, estimating the likelihood
of success is fundamental to the correct distribution of resources,
particularly in low-income environments.

@@

@@

@@

\caption{Perceptions of likelihood of eradication}\label{fig:unnamed-chunk-4}
\end{figure}

@@

{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-10-1}

@@

\caption{Aggregated perceived likelihood of eradication over time, adjusted for sample selection and unadjusted.}\label{fig:unnamed-chunk-10}
\end{figure}

@@

Comments were also classified as descriptive or prescriptive. A majority
(59.26\%) were descriptive. Descriptive commenters were more pessimistic
(in regards to perceived years until eradication) than prescriptive
commenters, though this difference did not reach the level of
statistical significance (p = 0.21 using Pearson's Chi-squared test).
Descriptive comments also received sentiment polarity scores which were
more negative than prescriptive comments, although again this difference
did not reach the level of statistical significance (p = 0.18).

@@

This study included the first authors of indexed journals. Though
certainly a group with important knowledge related to malaria, this
misses malaria control program employees, health agency workers, and
other stakeholders. Their experiences and viewpoints may be different
from those of academics, and arguably more relevant. Our de-biasing
method accounts for different response rates of ``senior'' vs.
``junior'' researchers, but does not take into account the fact that
first authors are generally more junior than senior authors (ie, the
pool from which we sampled may have been biased itself). To the extent
that in our results suggests that those with less experience (as
represented through publications) tended to be more ``optimistic''
regarding eradication, it is reasonable to assume that the restriction
of first authors may have lead to an overly optimistic sample, making
the results of the survey even more striking.

@@

The findings of our survey show researchers expressing hesitance about
the likelihood of eradication and suggesting a long time frame until it
is achieved. The causes for scepticism are diverse, but common themes
were the need for innovation, systemic challenges, and the complexity of
the disease and its transmission.

@@

\hypertarget{ref-Chen2018}{}
Chen, I., Cooney, R., Feachem, R.G.A., Lal, A., Mpanju-Shumbusho, W.,
2018. The lancet commission on malaria eradication. The Lancet 391,
1556--1558. \url{https://doi.org/10.1016/s0140-6736(18)30911-5}

@@
